### 2015-05-04 papers/histogram/thesis/thesis.tex
added tmmc and optimized ensemble to metheds section

\renewcommand{\title}{Optimizing Monte Carlo simulation of the
  square-well fluid} \renewcommand{\author}{Michael A. Perlin}
\newcommand{\institution}{Oregon State University}
\newcommand{\department}{Department of Physics}
\newcommand{\supervisor}{Dr. David Roundy} \renewcommand{\date}{21
  April 2015}

@@

\begin{algorithm}[tb]
  \caption{Metropolis Monte Carlo fluid simulation}
  \label{alg:MC}
  \begin{alg}

@@

  \item Randomly attempt to change the position of one sphere (in
    general, a single fluid ``atom'' or ``molecule'') to another
    location, rejecting the change if it results in a forbidden fluid
    configuration (e.g. two or more spheres overlap) and accepting the
    change otherwise. Attempting to move a single sphere is referred
    to as a move. \label{alg:mc_move}

  \item Repeat step \ref{alg:mc_move} for every other sphere in the
    fluid. Attempting to move every sphere once is referred to as an
    iteration of the simulation.
    \label{alg:mc_iteration}

  \item Repeat step \ref{alg:mc_iteration} indefinitely, or until data
    of sufficient quality has been generated, periodically collecting
    and dumping statistics on fluid states (e.g. energy, pair
    distribution histograms, etc.) to data files.

@@

Algorithm \ref{alg:MC} provides a sketch of unbiased, Metropolis Monte
Carlo fluid simulations. Such an algorithm is ``unbiased'' in the
sense that it collects statistics (i.e. data) on all valid system
configurations with equal probability. Statistics whose collection
time scales as $\mathcal O\p{1}$ with system size, meaning that
increasing the number of spheres $N$ in the simulation does not affect
collection time, can be collected after every move (defined in the
algorithm). Statistics whose collection time scales as $\mathcal
O\p{N}$, meaning that doubling $N$ doubles collection time, can be
collected after each iteration. In general, collection with $\mathcal
O\sp{\chi\p{N}}$ time scaling should not occur more often than once
every $\chi\p{N}$ moves (where $\chi\p{N}$ may be an arbitrary
function of $N$ void of constant prefactors, e.g. $N\log N$, or
$2^N$). These collection rules ensure that scaling up simulations does
not cause them to asymptotically spend all computation time collecting
statistics, and no time actually simulating the fluid. Collected
statistics are used to find thermodynamic properties of the simulated
fluid.

@@

The sort of Monte Carlo simulations described above are only capable
of collecting statistics on functions of system microstates, and
finding correlations between these functions. For example, in
simulation one might periodically compute both the energy $E\p{s}$ and
some system property $X\p{s}$, both of which are determined by the
microstate $s$, in order to find the mean value of $X$ at any given
energy $E$, i.e. $\bk{X}_E$. In the real world, however, information
about a system's microstates, and by extension functions of
microstates (e.g. the energy $E$), is inaccessible. One therefore
cannot measure $\bk{X}_E$ directly. Instead, one typically measures
the dependence of thermodynamic properties on macroscopic state
variables such as temperature, i.e. $\bk{X}_T$.

@@

Due to the fact that Metropolis Monte Carlo simulations sample all of
state space randomly and without preference, a histogram $H\p{X}$ of
observations of some system property $X$ is directly proportional to
the density of states $D\p{X}$; that is, the number of observations
$H\p{X_0}$ of the system with $X\p{s}=X_0$ is proportional to the
total number of states $s$ for which $X\p{s}=X_0$. It is sometimes the
case, however, that the density of states in some region $R$ in the
range of possible $X$ is so low that it is practically impossible (via
Metropolis Monte Carlo) to sufficiently sample $R$, that is,
accumulate a statistically significant histogram $H\p{X\in R}$, in any
reasonable amount of time.

@@

Histogram methods provide a means to address Metropolis Monte Carlo's
inability to sufficiently sample regions of parameter space with a low
density of states by introducing a bias into the otherwise random
sampling of state space. A weighting function $w$ is introduced, whose
domain is the value of some property $X\p{s}$ which depends on the
system state $s$. An additional condition is then added to step
\ref{alg:mc_move} of Algorithm \ref{alg:MC} in order to accept an
attempted move: the weights $w\sp{X\p{s_i}}$ and $w\sp{X\p{s_f}}$ of
the initial (pre-move) state $s_i$ and final (post-move) state $s_f$
are used to determine the probability $P\p{s_i\to s_f}$ of accepting
an otherwise valid move via
\begin{align}
  P\p{s_i\to s_f}=\min\set{\f{w\sp{X\p{s_f}}}{w\sp{X\p{s_i}}},1}.
  \label{eq:move_prob}
\end{align}
This formula means that when $w\sp{X\p{s_f}}>w\sp{X\p{s_i}}$, the move
$s_i\to s_f$ is accepted; when $w\sp{X\p{s_f}}<w\sp{X\p{s_i}}$, the
ratio of these weights determines the probability of accepting the
move. Due to the fact that only ratios of weights determine
$P\p{s_i\to s_f}$, the weights $w\sp{X\p{s}}$ are scale-invariant,
meaning that their effect on simulations is unchanged by scale factors
that are constant with respect to $X\p{s}$.

@@

  \item Repeat step \ref{alg:biased_mc_move} for every other sphere in
    the fluid.
    \label{alg:biased_mc_iteration}

  \item Repeat step \ref{alg:biased_mc_iteration} until data of
    sufficient quality has been generated.

@@

In this paper, we will consider weights $w\p{E}$ which depend only on
the energy $E$ of the system. Due to the fact that the square-well
fluid can only have discrete energies $E=-n\epsilon$, where $n$ is a
non-negative integer and $\epsilon$ is the well depth, in simulation
we store the weight function as an array of values. We may therefore
use the terms ``weight function,'' and ``weight array,'' and
``weights'' interchangeably.

@@

In our implementation of the simple flat method, we used $u=2$ and
$n_0=L\exp\p{\epsilon/kT_{\t{min}}}$, where $L$ is the number of
energy levels of the given square-well fluid and $T_{\t{min}}$ is a
minimum temperature of interest for the current simulation. The factor
of $L$ appears because initialization time should scale with the range
of energies of the simulated system. If we have twice the number of
energies, we should explore energy space for twice as long. To
estimate $L$, we multiplied the maximum number of spheres $M$ which
fit within a radius of $\lambda\sigma$ of a single sphere on a
face-centered cubic lattice (i.e. the maximum number of spheres with
which a given sphere could interact) by the total number of spheres
$N$, and divided the result by two so as to not double-count
interactions between spheres (so $L=MN/2$). The exponential
$\exp\p{\epsilon/kT_{\t{min}}}$, a Boltzmann factor for two adjacent
energies, appears because lower minimum temperatures of interest
warrant sampling lower energies, which in turn requires simulating for
longer (in step \ref{alg:simple_flat_sim} of Algorithm
\ref{alg:simple_flat}) because these energies have lower state
densities. The introduction of a minimum temperature may seem ad-hoc,
but all simulations in this paper require a choice of minimum
temperature anyways (addressed in Section \ref{sec:energy_range}). In
this paper, we will always use $kT_{\t{min}}=0.2\epsilon$.

@@

Unlike the previous histogram methods, the transition matrix Monte
Carlo (TMMC) method does not use weights, and merely compute the
density of states $D\p{E}$ of a system. The density of states can in
turn be used via the flat histogram method (Section
\ref{sec:flat_histogram}) to determine appropriate weights for
subsequent Monte Carlo simulations. TMMC introduces a new object: the
energy transition matrix, $T$, whose components $T_{ij}$ are the
probabilities that a system will transition {\it to} a state with
energy $E_i$ {\it from} a given state with energy $E_j$,
i.e. $T_{ij}=P\p{E_j\to E_i}$.

The transition matrix $T$ is trivially square and positive-definite,
but it is not symmetric. Given that Metropolis Monte Carlo samples all
of state-space randomly, meaning all possible system states are
equally likely to occur during simulation, a transition $E_i\to E_f$
from an energy with a low state density to one with a high state
density, $D\p{E_i}<D\p{E_j}$, is more probable than the inverse
transition $E_j\to E_i$ simply due to the fact that there are more
states with energy $E_i$ than those with energy $E_j$. Thus
$D\p{E_i}<D\p{E_j}$ implies $P\p{E_i\to E_j}>P\p{E_j\to E_i}$, which
means $T_{ji}>T_{ij}$. Said otherwise, the asymmetry of the transition
matrix is a direct consequence of the second law of thermodynamics:
systems are exceedingly likely to evolve towards macroscopic states
with higher accessible microstate densities.

In principle, computationally determining the transition matrix of a
system is straightforward. One need only simulate a system and collect
a histogram $\tilde T\p{E_i,E_j}$ of the energy transitions after each
move, directly after deciding whether to reject the move based on
whether it resulted in an (in)valid system state. That is, transition
data should be collected (only) on all moves as they would have been
accepted by Metropolis Monte Carlo, i.e. before deciding whether to
reject any given move for any reasons involving weights, or other
biasing schemes.

\vspace{3cm}

\fixme{schemes for initializing to collect transition matrix}

\fixme{normalizing transition matrix}

\fixme{computing density of states via transition matrix}

\fixme{storing transition matrix in the form $T\p{E,\Delta E}$}

\vspace{3cm}

For the square-well fluid, as with many systems, there is a maximum
amount $\Delta_ME$ by which a single move can change the energy of the
system. The transition matrix can therefore be stored in the form
$T\p{E,\Delta E}$

the transition matrix is band-diagonal because . In implementation,
the transition matrix can therefore be stored in the form
$T\p{E,\Delta E}$, making it an $L\times\Delta_ME$ matrix, where $L$
is the number of energy levels in a given system. Given that $L$
scales as $N\Delta_ME$, where $N$ is the number of spheres in the
square-well fluid, the memory required to store the transition matrix
therefore scales as $N\p{\Delta_ME}^2$, an improvement over the $N^2$
growth of a transition matrix in the form $T\p{E_i,E_j}$.

\fixme{Explain, provide algorithm. This method is sort of our own, but
  based on other work.}

\begin{algorithm}[H]
  \caption{Transition matrix Monte Carlo initialization}
  \label{alg:tmmc}
\end{algorithm}

@@

\fixme{Explain briefly. No need to provide algorithm.}

@@

\begin{algorithm}[H]
  \caption{Hybrid OE-TMMC simulation}
  \label{alg:oetmmc}
\end{algorithm}

>>>>

\renewcommand{\title}{Optimizing Monte Carlo simulation of the
  square-well fluid} \renewcommand{\author}{Michael A. Perlin}
\newcommand{\institution}{Oregon State University}
\newcommand{\department}{Department of Physics}
\newcommand{\supervisor}{Dr. David Roundy}
\renewcommand{\date}{08 May 2015}

@@

\begin{algorithm}[tb]
  \caption{Metropolis Monte Carlo fluid simulation}
  \label{alg:metropolis}
  \begin{alg}

@@

  \item Randomly attempt to change the position of one sphere (in
    general, a single fluid ``atom'' or ``molecule'') to another
    location, rejecting the change if it results in a forbidden fluid
    configuration (e.g. two or more spheres overlap) and accepting the
    change otherwise. Attempting to move a single sphere is referred
    to as a move. \label{alg:metropolis_move}

  \item Repeat step \ref{alg:metropolis_move} for every other sphere
    in the fluid. Attempting to move every sphere once is referred to
    as an iteration of the simulation.
    \label{alg:metropolis_iteration}

  \item Repeat step \ref{alg:metropolis_iteration} indefinitely, or
    until data of sufficient quality has been generated, periodically
    collecting and dumping statistics on fluid states (e.g. energy,
    pair distribution histograms, etc.) to data files.

@@

Algorithm \ref{alg:metropolis} provides a sketch of unbiased,
Metropolis Monte Carlo fluid simulations. Such an algorithm is
``unbiased'' in the sense that it collects statistics (i.e. data) on
all valid system configurations with equal probability. Statistics
whose collection time scales as $\mathcal O\p{1}$ with system size,
meaning that increasing the number of spheres $N$ in the simulation
does not affect collection time, can be collected after every move
(defined in the algorithm). Statistics whose collection time scales as
$\mathcal O\p{N}$, meaning that doubling $N$ doubles collection time,
can be collected after each iteration. In general, collection with
$\mathcal O\sp{\chi\p{N}}$ time scaling should not occur more often
than once every $\chi\p{N}$ moves, where $\chi\p{N}$ may be an
arbitrary function of $N$ void of constant prefactors, e.g. $N\log N$,
or $2^N$. These collection rules ensure that scaling up simulations
does not cause them to asymptotically spend all computation time only
collecting statistics, or only simulating the fluid. Collected
statistics are used to find thermodynamic properties of the simulated
fluid.

@@

The sort of Monte Carlo simulations described above are only capable
of collecting statistics on functions of system microstates, and
finding correlations between these functions. For example, in
simulation one might periodically compute both the energy $E\p{s}$ and
some system property $X\p{s}$, both of which are determined by the
microstate $s$, in order to find the mean value of $X$ at any given
energy $E$, i.e. $\bk{X}_E$. In the real world, however, information
about a system's microstates, and by extension information about
functions of microstates (e.g. the energy $E$), is inaccessible. One
therefore cannot measure $\bk{X}_E$ directly. Instead, one typically
measures the dependence of thermodynamic properties on macroscopic
state variables such as temperature, i.e. $\bk{X}_T$.

@@

Due to the fact that Metropolis Monte Carlo simulations sample all of
state space randomly and without preference, a histogram $H\p{X}$ of
observations of some system property $X$ is directly proportional to
the density of states $D\p{X}$; that is, the number of observations
$H\p{X_0}$ of the system with $X\p{s}=X_0$ is proportional to the
total number of states $s$ for which $X\p{s}=X_0$. It is sometimes the
case, however, that the density of states in some region $R$ in the
range of possible $X$ is so low that it is practically impossible (via
Metropolis Monte Carlo) to sufficiently sample $R$, that is, to
accumulate a statistically significant histogram $H\p{X\in R}$, in any
reasonable amount of time. Given that the entropy $S\p{X}$ can be
expressed in terms of the number of microstates $\Omega\p{X}$ as
$S\p{X}=k\ln\Omega\p{X}\propto\ln D\p{X}$, we may refer to regions $R$
low state densities $D\p{X\in R}$ as low entropy states.

@@

Histogram methods provide a means to address Metropolis Monte Carlo's
inability to sufficiently sample low entropy states by introducing a
bias into the otherwise random sampling of state space. A weighting
function $w$ is introduced, whose domain is the value of some property
$X\p{s}$ which depends on the system state $s$. An additional
condition is then added to step \ref{alg:metropolis_move} of Algorithm
\ref{alg:metropolis} in order to accept an attempted move: the weights
$w\sp{X\p{s_i}}$ and $w\sp{X\p{s_f}}$ of the initial (pre-move) state
$s_i$ and final (post-move) state $s_f$ are used to determine the
probability $P_m\p{s_i\to s_f}$ of accepting an otherwise valid move
via
\begin{align}
  P_m\p{s_i\to s_f}=\min\set{\f{w\sp{X\p{s_f}}}{w\sp{X\p{s_i}}},1}.
  \label{eq:move_prob}
\end{align}
This formula means that when $w\sp{X\p{s_f}}>w\sp{X\p{s_i}}$, the move
$s_i\to s_f$ is accepted; when $w\sp{X\p{s_f}}<w\sp{X\p{s_i}}$, the
ratio of these weights determines the probability of accepting the
move. Due to the fact that only ratios of weights determine
$P_m\p{s_i\to s_f}$, the weights $w\sp{X\p{s}}$ are scale-invariant,
meaning that their effect on simulations is unchanged by scale factors
that are constant with respect to $X\p{s}$.

@@

  \item Repeat step \ref{alg:biased_mc_move} the simulation produces
    statistics of sufficient quality.

@@

In this paper, we will consider weights $w\p{E}$ which depend only on
the energy $E$ of the square-well fluid. Due to the fact that the
square-well fluid can only have discrete energies $E=-n\epsilon$,
where $n$ is a non-negative integer and $\epsilon$ is the well depth,
in simulation we store the weight function as an array of values. We
may therefore use the terms ``weight function,'' and ``weight array,''
and ``weights'' interchangeably.

@@

In our implementation of the simple flat method, we used $u=2$ and
$n_0=L\exp\p{\epsilon/kT_{\t{min}}}$, where $L$ is the number of
energy levels of the given square-well fluid and $T_{\t{min}}$ is a
minimum temperature of interest for the current simulation. The factor
of $L$ appears because initialization time should scale with the range
of energies of the simulated system. If we have twice the number of
energies, we should explore energy space for twice as long. To
estimate $L$, we multiplied the maximum number of spheres $M$ which
fit within a radius of $\lambda\sigma$ of a single sphere on a
face-centered cubic lattice (i.e. the maximum number of spheres with
which a given sphere could interact) by the total number of spheres
$N$, and divided the result by two so as to not double-count
interactions between spheres (so $L=MN/2$). The exponential
$\exp\p{\epsilon/kT_{\t{min}}}$, a Boltzmann factor for two adjacent
energies, appears because lower minimum temperatures of interest
warrant sampling lower energies, which in turn requires simulating for
longer (in step \ref{alg:simple_flat_sim} of Algorithm
\ref{alg:simple_flat}) because these energies have lower state
densities. The introduction of a minimum temperature may seem ad-hoc,
but all simulations in this paper require a choice of minimum
temperature anyways (discussed in Section \ref{sec:energy_range}). In
this paper, we will always use $kT_{\t{min}}=0.2\epsilon$.

@@

\fixme{add citations}

Unlike the previous histogram methods, the transition matrix Monte
Carlo (TMMC) method does not use weights, and merely compute the
density of states $D\p{E}$ of a system\cite{tmmc}. The density of
states can in turn be used to determine flat histogram weights via
(\ref{eq:flat_weights}). TMMC introduces a new object: the energy
transition matrix, $T$, whose components $T_{ij}$ are the
probabilities that a system will transition {\it to} a state with
energy $E_i$ {\it from} a given state with energy $E_j$,
i.e. $T_{ij}=P\p{E_j\to E_i}$.

The transition matrix is trivially square and positive-definite, but
it is not symmetric. Given that Metropolis Monte Carlo samples all of
state-space randomly, meaning all possible system states are equally
likely to occur during simulation, a transition $E_i\to E_f$ from an
energy with a low state density to one with a high state density,
$D\p{E_i}<D\p{E_j}$, is more probable than the inverse transition
$E_j\to E_i$ simply due to the fact that there are more states with
energy $E_i$ than those with energy $E_j$. Thus $D\p{E_i}<D\p{E_j}$
implies $P\p{E_i\to E_j}>P\p{E_j\to E_i}$, which means
$T_{ji}>T_{ij}$. The asymmetry of the transition matrix is precisely
the origin of the second law of thermodynamics: systems are
exceedingly likely to evolve towards macroscopic states with higher
accessible microstate densities.

Knowing the transition matrix of a system determines, among other
properties, the system's density of states. Consider an ensemble of
Metropolis Monte Carlo simulations with a distribution of energies at
a time $t$ given by $\hat D^{t}\p{E}$, such that the probability of
any given simulation to have an energy $E_i$ at a time $t$ is $\hat
D_i^{t}=\hat D^{t}\p{E_i}$. The probability $\hat D_i^{t}$ can be
expressed in terms of the distribution $\hat D^{t-1}\p{E}$ one time
step (i.e. one Metropolis Monte Carlo move) prior by
\begin{align}
  \hat D_i^{t}=\sum_jP\p{E_i\to E_j}\hat D_j^{t-1} =\sum_j T_{ij}\hat
  D_j^{t-1}.
  \label{eq:transition_evolution}
\end{align}
At equilibrium, the probability distributions $\hat D^{t}=\hat
D^{t-1}$. Furthermore, the distribution of states in an equilibrium
ensemble of Metropolis Monte Carlo simulations is precisely the
density of states $D\p{E}$\fixme{explain why?}, which means
$D_i=T_{ij}D_j$. Simultaneously expressing this condition for all
energies,
\begin{align}
  D=TD. \label{eq:dos_eigen}
\end{align}
Finding the density of states of a system can thus be reduced to
finding the eigenvector of the transition matrix which has a
corresponding unit eigenvalue\fixme{why is this eigenvector
  unique?}. Once one has determined the density of states of a system,
one can initialize weights via the flat histogram method,
i.e. (\ref{eq:flat_weights}).

Computationally determining the transition matrix of a system is
straightforward: one need only simulate the system and collect a
histogram $\tilde T_{ij}=\tilde T\p{E_i,E_j}$ of the energy
transitions after each Metropolis Monte Carlo move, i.e. after step
\ref{alg:metropolis_move} of Algorithm \ref{alg:metropolis}. A
histogram of transitions $\tilde T\p{E_i,\bar E_j}_{}$ from a fixed
energy $\bar E_j$ to energies $E_i$ is then proportional to the
probability distribution of transitions from $\bar E_j$, i.e. the
transition matrix $T\p{E_i,\bar E_j}$. Crucially, this proportionality
does not depend on the history of a simulation, or on how the
simulation got to be in the energy $\bar E_j$, so long as samples of
the energy $\bar E_j$ are themselves void of systemic
bias. Furthermore, one does not actually need to transition from $E_j$
to $E_i$ in order to collect statistics on $T\p{E_i,E_j}$: after
deciding whether to accept a move based on whether it results in a
valid system state, one can increment $\tilde T\p{E_i,E_j}$ before
deciding to reject the move for some other reason, e.g. because
$w\p{E_j}\gg w\p{E_i}$, or even because one wishes to collect more
statistics on $T\p{E_i,\bar E_j}$.

The proportionality $\tilde T_{ij}\propto T_{ij}$ can be made explicit
by enforcing that the probabilities of all transitions from an energy
$E_j$ sum to unity, i.e.
\begin{align}
  \sum_iP\p{E_j\to E_i}=\sum_iT_{ij}=1,
  \label{eq:transition_norm_condition}
\end{align}
which implies
\begin{align}
  T_{ij}=\f{\tilde T_{ij}}{\sum_i\tilde T_{ij}}.
  \label{eq:transition_normalization}
\end{align}
To compute $T$ from $\tilde T$, one must therefore normalize each row
of $\tilde T$ independently.

If a system has $L$ energy levels, the transition histogram $\tilde
T\p{E_i,E_j}$ is an $L\times L$ matrix. Given that $L$ is proportional
to the system size $N$, the memory footprint of $\tilde T\p{E_i,E_j}$
grows as $N^2$. For most systems, however, there will be a maximum
energy $M$ independent of system size by which a single move can
change the energy of the system. The transition histogram for such
systems is therefore band-diagonal, which one can exploit to save
memory. In particular, it is natural to store the transition histogram
in the form $\tilde T_d\p{E,\Delta E}$, which gives the probability
that a system will transition from the energy $E$ to the energy
$E+\Delta E$ in a single move. One can convert between $\tilde
T_d\p{E,\Delta E}$ and $\tilde T\p{E_i,E_j}$ using the relation
\begin{align}
  \tilde T\p{E_i,E_j}=\tilde T_d\p{E_j,E_i-E_i},
\end{align}
which means, from (\ref{eq:transition_normalization}),
\begin{align}
  T_{ij}=\f{\tilde T_d\p{E_j,E_i-E_j}}{\sum_i\tilde T_d\p{E_j,E_i-E_j}}
  =\f{\tilde T_d\p{E_j,E_i-E_j}}
  {\sum_{\Delta E}\tilde T_d\p{E_j,\Delta E}}.
  \label{eq:transition_conversion}
\end{align}
Expressed in this form, the memory footprint of the transition
histogram grows linearly with system size $N$ and the maximal energy
difference $M$.

While one could imagine many initialization routines to determine the
transition matrix $T$, we use Algorithm \ref{alg:tmmc}, which is a
modified version of a routine provided in \cite{tmmc_prl}\fixme{is
  this citation correct?}. The primary motivation behind this
algorithm is such: as lower energies are always more difficult to
sample than higher energies, transitions to lower energies should
always be accepted, while transitions to higher energies should pass a
probabilistic acceptance test. To understand the probability in step
\ref{alg:tmmc_prob} of Algorithm \ref{alg:tmmc} of accepting a move
$s_i\to s_f$ for which $E_f>E_i$, first consider the case of $c=0$ and
$T_{\t{min}}\to0$, in which case the probability becomes
\begin{align}
  P_m\p{s_i\to s_f}=\f{\p{\tilde T_d\p{E_i,\Delta E}}\big/
    \p{\sum_{\Delta E'}\tilde T_d\p{E_i,\Delta E'}}}
  {\p{\tilde T_d\p{E_f,-\Delta E}}\big/
    \p{\sum_{\Delta E'}\tilde T_d\p{E_f,\Delta E'}}}
  =\f{T\p{E_i,\Delta E}}{T\p{E_f,-\Delta E}}
  =\f{P\p{E_i\to E_f}}{P\p{E_f\to E_i}},
  \label{eq:tmmc_prob_simplified}
\end{align}
where $\Delta E=E_f-E_i$ and $P\p{E_m\to E_n}$ is the unbiased
probability of a transition from an energy $E_m$ to an energy
$E_n$. The result in (\ref{eq:tmmc_prob_simplified}) is the
probability of accepting moves $s_i\to s_f$ required for simulation
via Algorithm \ref{alg:tmmc} to yield a flat energy
histogram\cite{tmmc}. Unlike the previous histogram methods, we desire
a flat histogram during initialization via Algorithm \ref{alg:tmmc}
not because we are constructing weights $w\p{E}$, but because we wish
to sample the transition histogram $\tilde T_d\p{E,\Delta E}$ equally
at all energies $E$. \fixme{briefly explain
  (\ref{eq:tmmc_prob_simplified}) by considering a two-level system,
  and extending to larger systems}.

Aside from a minimum temperature $T_{\t{min}}$ and an end condition,
Algorithm \ref{alg:tmmc} contains one free parameter, $c$, used in
step \ref{alg:tmmc_prob} to modify (\ref{eq:tmmc_prob_simplified}).
This parameter makes the rejection of transitions more conservative
when there are too few statistics in the histogram $\tilde T_d$ to be
confident in the current estimate of the transition matrix. In order
to make the effects of $c$ negligible as $\tilde T_d$ collects more
statistics throughout initialization, $c$ should be of order unity,
though its optimal value is generally system-dependent. In our
implementation of Algorithm \ref{alg:tmmc}, we use $c=16$.

The last interesting part of the probability in step
\ref{alg:tmmc_prob} of Algorithm \ref{alg:tmmc} is the limit on
$P_m\p{s_i\to s_f}$ to a minimum value of $\exp\p{-\Delta
  E/kT_{\t{min}}}$. This limit caps the bias on energies to that
introduced by canonical (fixed-temperature) weights at a temperature
of $T_{\t{min}}$. As a result, Algorithm \ref{alg:tmmc} will not waste
time oversampling energies which are unimportant for determining
system properties at temperatures $T\ge T_{\t{min}}$.

\begin{algorithm}[tb]
  \caption{Transition matrix Monte Carlo initialization}
  \label{alg:tmmc}
  \begin{alg}

  \item Initialize a transition histogram $\tilde T_d\p{E,\Delta E}=0$
    for all $E$ and $\Delta E\in\sp{-M,M}$, where $M$ is the possible
    maximum energy transition of a single move.

  \item Randomly attempt to change the position of one sphere,
    tentatively accepting the transition from state $s_i$ to state
    $s_f$ if $s_f$ is not a forbidden fluid configuration.
    \label{alg:tmmc_move}

  \item Let $\Delta E=E_f-E_i$ and increment $\tilde
    T_d\p{E_i,\Delta{E}}$, where $E_k$ is the energy of state $s_k$.

  \item If $\Delta E<0$, accept the move. Otherwise
    \begin{enumerate}
    \item compute the normalization factor
      \begin{align*}
        n_k=\sum_{\Delta E'}\tilde T_d\p{E_k,\Delta E'}
      \end{align*}
      for $k\in\set{i,f}$, and

    \item accept the move $s_i\to s_f$ with probability
      \begin{align*}
        P_m\p{s_i\to s_f}
        =\max\set{\f{\p{\tilde T_d\p{E_i,\Delta E}+c}/\p{n_i+c}}
          {\p{\tilde T_d\p{E_f,-\Delta E}+c}/\p{n_f+c}},
          \exp\p{-\f{\Delta E}{kT_{\t{min}}}}}
      \end{align*}
      for a predetermined number $c$ and temperature $T_{\t{min}}$.
      \label{alg:tmmc_prob}
    \end{enumerate}

  \item If some predetermined initialization end condition has
    \emph{not} been met, return to step \ref{alg:tmmc_move}.

@@

All histogram methods introduced in this paper thus far have focused
on determining a weight array $w\p{E}$ for a flat energy histogram
$H\p{E}$, in order to sample and collect statistics on all energies
equally. This motivation considers the quantity of statistics, but not
their quality. Biased simulations collect statistics on low energy
states by first getting into these states, and then rejecting moves
which move the system into higher energy states. Statistics on low
energy states will therefore generally be highly correlated, as they
are based on many samples of only a few low energy regions of a
system's energy landscape. The optimized ensemble
(OE)\cite{optimized_ensemble} attempts to address the autocorrelation
between low energy samples by finding weights to maximize the rate at
which a simulation makes round trips between low and high energy
states. Maximizing the round trip rate, in turn, maximizes the rate at
which a simulation makes independent, uncorrelated samples of low
energies.

The optimized ensemble works by considering an ensemble of
simulations, each of which defines the position of a ``walker'' in
energy space. Given that we want walkers to move back and forth
between two extrema $E_+$ and $E_-$ of some specified energy range, we
label each walker by which extrema it has visited most recently. We
can then identify ``down-going'' (``up-going'') walkers by those which
more recently visited $E_+$ ($E_-$), which means that to make a round
trip they first needed to get to $E_-$ ($E_+$). Denoting the total
walker and down-going density at an energy $E$ respectively by
$n\p{E}$ and $n_+\p{E}$, the ratio $f\p{E}=n_+\p{E}/n\p{E}$ gives the
proportion of walkers at $E$ which are down-going. We also denote the
net and down-going walker diffusivity at an energy $E$ respectively by
$\alpha\p{E}$ and $\alpha_+\p{E}$.

We can meaningfully discuss a down-going walker flux $j\p{E}$, which
is defined by
\begin{align}
  j=\f{dn_+}{dt}=\alpha_+\f{dn_+}{dE}=f\alpha\f{dn_+}{dE}.
  \label{eq:walker_current_definition}
\end{align}
Computing
\begin{align}
  n\f{df}{dE}=n\sp{\f1{n}\f{dn_+}{dE}-\f{n_+}{n^2}\f{dn}{dE}}
  =\f{dn_+}{dE}-f\f{dn}{dE},
  \label{eq:walker_intermediate}
\end{align}
we can say that
\begin{align}
  \f{dn_+}{dE}=n\f{df}{dE}+f\f{dn}{dE}.
  \label{eq:walker_gradient}
\end{align}
Substituting (\ref{eq:walker_gradient}) into
(\ref{eq:walker_current_definition}) yields
\begin{align}
  j=f\alpha\sp{n\f{df}{dE}+f\f{dn}{dE}}.
  \label{eq:walker_current_full}
\end{align}
If the energy histogram does vary strongly with energy, e.g. because
we are simulating with flat histogram weights, then $dn/dE\approx0$,
which means
\begin{align}
  j=f\alpha n\f{df}{dE}.
  \label{eq:walker_current}
\end{align}
To formalize the approximation made between
(\ref{eq:walker_current_full}) and (\ref{eq:walker_current}), we can
observe that
\begin{align}
  \f{f~dn/dE}{n~df/dE}
  =\p{\f{dn_+/dE-fdn/dE}{f~dn/dE}}^{-1}=\p{\f1f\f{dn_+}{dn}-1}^{-1}
  =\p{\f{d\ln n_+}{d\ln n}-1}^{-1}.
  \label{eq:walker_current_intermediate_ratio}
\end{align}
For (\ref{eq:walker_current}) to be valid, we want the ratio in
(\ref{eq:walker_current_intermediate_ratio}) to vanish. If the walker
density $n$ varies weakly with energy, e.g. because we are simulating
with flat histogram weights, then $\ln n$ varies even weaker
still. The down-going walker density $n_+$, meanwhile must run from
$n\p{E_-}=0$ to $n\p{E_+}=n\p{E_+}$ meaning $\ln n_+$ runs from $\ln
n\p{E_-}\to-\infty$ to a finite positive value $\ln n_+\p{E_+}=\ln
n\p{E_+}$. It is therefore reasonable to assume that $d\ln n_+/d\ln
n\gg 1$, which makes the ratio in
\ref{eq:walker_current_intermediate_ratio} vanish.

We can rearrange (\ref{eq:walker_current}) and integrate from the low
energy state $E_-$, at which $f\p{E_-}=0$, to the high energy state
$E_+$, at which $f\p{E_+}=1$, to say
\begin{align}
  \int_{f=0}^1\f{df}{j}=\int_{E=E_-}^{E_+}\f{dE}{f\alpha n}.
  \label{eq:walker_current_setup}
\end{align}
At equilibrium, $j=dn_+/dt$ is constant, which means we can evaluate
one of the integrals in (\ref{eq:walker_current_setup})
\begin{align}
  \f1j=\int_{E=E_-}^{E_+}\f{dE}{f\alpha n}.
  \label{eq:walker_current_integral}
\end{align}

The optimized ensemble method minimizes the integral in
(\ref{eq:walker_current_integral}), thereby maximizing the down-going
walker current $j$, by varying the walker density $n$ with the
constraint that $n$ must remain normalized. With the additional
assumption that $\alpha\p{E}$ does not strongly depend on the weights
$w\p{E}$, the authors of \cite{optimized_ensemble} find that the
optimal walker density $n$ is
\begin{align}
  n_{\t{opt}}=\f1{\sqrt{\lambda\alpha}},
  \label{eq:optimal_walker_density}
\end{align}
where $\lambda$ is the Lagrange multiplier used in the minimization
process, and here acts merely as a normalization factor.

Equating the set of observations in a single simulation with a single
observation of an ensemble of simulations, the walker density $n\p{E}$
is proportional to the energy histogram $H\p{E}=\tilde D\p{E}w\p{E}$,
which means
\begin{align}
  w_{\t{opt}}=\f1{D\sqrt{\alpha}},
  \label{eq:optimized_weights}
\end{align}
where we neglect the distinction between $\tilde D$ and $D$, which has
no effect on the weights $w$.

The algorithm provided in \cite{optimized_ensemble} to determine and
construct weights given by \ref{eq:optimized_weights} involves
simulating with and iteratively modifying flat histogram weights. This
algorithm is not meant to construct weights from scratch, but rather
to optimize them. We therefore will not be comparing the optimized
ensemble to other methods. We do, however, credit this method for
motivating the hybrid OE-TMMC method in Section \ref{sec:oetmmc}.

@@

\begin{algorithm}[H]
  \caption{Modifying transition weights via the optimized ensemble}
  \label{alg:oetmmc}
\end{algorithm}
